A weakly structured stem for human origins in Africa

### 2022-09-09 paper.tex
pushed Middle Stone Age changes

We quantify migration rates among populations after their divergence $\approx120ka$.
Migration, prior to the Bantu expansion, between the ancestors of the Nama and other groups
is an order of magnitude weaker than that observed between between western and eastern Africans (Table \ref{tab:migration-rates}).
All models infer relatively high gene flow between Ethiopia and western Africa
($m\approx2\times10^{-4},$ the constant proportion of immigrant lineages per generation over 60ka).
We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations \citep{Molinaro2019-rf},
comprising over half of their genetic ancestry, estimated to be 64--65\%.
We observe significant gene flow from the Amhara and Oromo into the Nama, a signal which is likely a proxy for the movement of
eastern African caprid and cattle pastoralists \citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. While this gene flow is not apparent from the ADMIXTURE plot (Figure~\ref{fig:diversity}),
the ancestry is likely grouped into the Khoe-San component which has drifted appreciably from its ancestral eastern African source.
Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions suggested by \texttt{ADMIXTURE}
(Figure~\ref{fig:diversity}).

@@

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained (standard error (SE) $=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
479ka. The timing of this divergence was also poorly constrained (SE $=166$ka).
These populations evolve independently until 119ka
(SE $=28$ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the western Africans and
eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25ka (SE $=0.6$ka).
The later Stem 2 contribution to the western African Mende resulted in
better model fits ($\Delta LL \approx 60,000$).
Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
central Africa, broadly speaking. The differing proportions in the Nama and
eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

>>>>

We quantify migration rates among populations after their divergence $\approx120ka$.
Migration, prior to the Bantu expansion, between the ancestors of the Nama and other groups  
is an order of magnitude weaker than that observed between between western and eastern Africans (Table \ref{tab:migration-rates}).
All models infer relatively high gene flow between Ethiopia and western Africa
($m\approx2\times10^{-4},$ the constant proportion of immigrant lineages per generation over 60ka).
We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. 
We observe significant gene flow from the Amhara and Oromo into the Nama, a signal which is likely a proxy for the movement of
eastern African caprid and cattle pastoralists \citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. While this gene flow is not apparent from the ADMIXTURE plot (Figure~\ref{fig:diversity}), 
the ancestry is likely grouped into the Khoe-San component which has drifted appreciably from its ancestral Eastern African source. 
Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions suggested by \texttt{ADMIXTURE}
(Figure~\ref{fig:diversity}).

@@

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained (standard error (SE) $=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
479ka. The timing of this divergence was also poorly constrained (SE $=166$ka).
These populations evolve independently until 119ka
(SE $=28$ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the western Africans and
eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25ka (SE $=0.6$ka).
The later Stem 2 contribution to the western African Mende resulted in
better model fits ($\Delta LL \approx 60,000$).
Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
Central Africa, broadly speaking. The differing proportions in the Nama and
eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.
